---
title: "Lab 06 — Hypothesis Testing from Claims"
author: "EAES 480 — Dylan Enwia Modern Statistics in Earth & Environmental Science"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    df_print: paged
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

# Overview

This lab introduces **hypothesis testing** as an extension of sampling, bootstrap distributions, and uncertainty quantification.

The core idea is simple:

> A **claim** about the population becomes a **null hypothesis**.

You will use the **US-AMS simplified AmeriFlux dataset** again and test claims about population means using:

- one sample,
- a bootstrap estimate of uncertainty,
- a z-statistic,
- and a p-value.

This lab emphasizes **reasoning from claims to tests**, not just mechanics.

---

# Learning goals

By the end of this lab, you should be able to:

- Turn a verbal claim into a null hypothesis
- Choose a variable and define the relevant population parameter
- Draw a sample and estimate a mean
- Use bootstrap resampling to estimate uncertainty in that mean
- Compute a z-statistic and p-value
- Interpret a hypothesis test in plain language
- Connect confidence intervals and hypothesis tests

---

# Data

## Load and inspect

```{r load_data, echo=TRUE, eval=TRUE}
# GOAL: Read the dataset and clean the names.

df <- read_csv("data/us-ams-simple.csv", na = c("-9999")) %>%
  clean_names() %>%
  filter(year_local == 2023)

glimpse(df)
```

## Derive time columns (if needed)

```{r derive_time, echo=TRUE, eval=TRUE}
# GOAL: Create date / month / day_night columns as in earlier labs.

df <- df %>%
  mutate(
    date = as.Date(floor(doy) - 1, origin = paste0(year_local, "-01-01")),
    month = month(date, label = TRUE, abbr = TRUE),
    day_night = if_else(daytime == 1, "Day", "Night")
  )
```

---

# Part I — From claim to hypothesis test

## 1) Choose a variable

Choose **one numeric variable** from the dataset.

Possible examples:
- `swc`
- `ta`
- `fc`
- `gpp`
- `le`

```{r choose_variable, echo=TRUE, eval=TRUE}
target_var <- "ta"

# CHECK: does this variable exist?
target_var %in% names(df) 
```

## 2) Describe the population parameter

Write the population parameter you care about:

> $\mu$ = the true 2023 population mean of `ta`

Fill in the blank in words:

> In this lab, my population parameter is the true annual mean of **`air temperature (ta)`** at the US-AMS site.

---

## 3) Make a claim

Write a claim about the population mean for your variable.

Examples:
- “The annual mean soil water content is 0.30.”
- “The annual mean air temperature is 10 °C.”
- “The annual mean latent heat flux is 40 W m$^{-2}$.”

Now turn your claim into hypotheses:

- Null hypothesis: $H_0: \mu = \mu_0$
- Alternative hypothesis: $H_A: \mu \neq \mu_0$

Fill these in:

> My claim is: `The annual mean of air temperature is 12 °C.`

> $H_0$: `The true annual mean of air temperature is 12 °C.`

> $H_A$: `The true annual mean of air temperature will not be 12 °C.`

---

## 4) Draw one sample and compute the sample mean

```{r draw_sample, echo=TRUE, eval=TRUE}
# GOAL: Draw one random sample and compute its mean.

set.seed(10)

n_samp <- 100   # e.g., 200

samp <- df %>%
  filter(!is.na(.data[[target_var]])) %>%
  slice_sample(n = n_samp)

sample_mean <- mean(samp[[target_var]], na.rm = TRUE)

sample_mean
```

**Prompt (2–3 sentences):** How does your sample mean compare to your claim? Is it close, or already quite different?

> *It doesn't vary by much it is pretty close. My claim was that air temperature would be 12 °C and for my sample mean its about 13.3 °C.*

---

## 5) Bootstrap the sample mean

```{r bootstrap_means, echo=TRUE, eval=TRUE}
# GOAL: Bootstrap the sample mean to estimate uncertainty.

set.seed(100)

B <- 2000   # e.g., 4000

boot_means <- replicate(
  B,
  mean(sample(samp[[target_var]], size = n_samp, replace = TRUE), na.rm = TRUE)
)

head(boot_means)
```

## 6) Estimate uncertainty

```{r bootstrap_summary, echo=TRUE, eval=TRUE}
# GOAL: Compute the bootstrap SE and 95% CI.

boot_se <- sd(boot_means)
boot_ci <- quantile(boot_means, probs = c(0.025, 0.095))

tibble(
  sample_mean = sample_mean,
  boot_se = boot_se,
  ci_lower = boot_ci[1],
  ci_upper = boot_ci[2]
)
```

**Prompt (2–3 sentences):** What does the bootstrap SE tell you? What does the 95% CI tell you?

> *The bootstrap SE shows the uncertainty in the estimate by the sample mean varying at repeated samples. My boot_se is 0.96 which shows that the sample mean has uncertainty. The 95% CI tells us that the range for the true 2023 annual mean of air temperature is between 11.5 to 12.1* 

---

## 7) Convert the claim into a formal test

Define the null value from your claim.

```{r hypothesis_test, echo=TRUE, eval=TRUE}
# GOAL: Compute z and p-value for your claim.

mu0 <- 12   # numeric claim value

z_stat <- (sample_mean - mu0) / boot_se

p_value <- 2 * pnorm(-abs(z_stat))

tibble(
  sample_mean = sample_mean,
  mu0 = mu0,
  boot_se = boot_se,
  z_stat = z_stat,
  p_value = p_value
)
```

---

## 8) Interpret the result

Use your p-value and CI to answer:

- Does the 95% CI include the null value?
- Is the p-value small or large?
- Would you reject the null hypothesis at $\alpha = 0.05$?

Fill in:

> The null value `is` inside my 95% CI.

> My p-value is `0.169`, which is large.

> Therefore, I would not reject $H_0$ at the 0.05 level.

**Prompt (4–6 sentences):** Explain your decision in plain language. Do the data make the claim look plausible or surprising? Be careful: “fail to reject” does not mean “prove true.”

> *Basically the data here looks plausible. The reason I say this is first my p-value is large and my null value of 12 is inside my 95% CI. From this I wouldn't reject the $H_0$ at the 0.05 level. Overall the data doesn't provide strong evidence that the true value is different from 12. But failing to reject the null hypothesis doesn't prove that the true value is exactly 12, but it means that the data isn't surprising if the true value was 12. *

---

## 9) Visualize the hypothesis test

```{r plot_test, echo=TRUE, eval=TRUE}
# GOAL: Plot the bootstrap distribution, sample mean, CI, and null value.

ggplot(tibble(boot_mean = boot_means), aes(x = boot_mean)) +
  geom_histogram(bins = 50, alpha = 0.8) +
  geom_vline(xintercept = sample_mean, linetype = "dashed", linewidth = 1.1) +
  geom_vline(xintercept = boot_ci, color = "red", linewidth = 1.1) +
  geom_vline(xintercept = mu0, color = "blue", linewidth = 1.1) +
  theme_classic(base_size = 18) +
  labs(
    x = paste0("Bootstrap mean of ", target_var),
    y = "Count",
    title = "Bootstrap distribution and null claim",
    subtitle = "Dashed = sample mean; Red = 95% CI; Blue = null value"
  )
```

**Prompt (2–3 sentences):** Does the plot agree with your p-value-based conclusion? Explain.

> *Yes, the plot agrees with the p-value–based conclusion. The blue line in the plot is inside the 95% CI and within the bootstrap distribution range. This shows that the value of 12 here is plausible and isn't surprising thus matches the p-value of 0.169 and supports the decision to fail to reject $H_0$.*

---

## 10) Try a second claim (short extension)

Keep the same variable, but test a **different null value**.

```{r second_claim, echo=TRUE, eval=TRUE}
# GOAL: Test a second claim using the same sample and bootstrap SE.

mu0_alt <- 10

z_stat_alt <- (sample_mean - mu0_alt) / boot_se
p_value_alt <- 2 * pnorm(-abs(z_stat_alt))

tibble(
  mu0_alt = mu0_alt,
  z_stat_alt = z_stat_alt,
  p_value_alt = p_value_alt
)
```

**Prompt (3–4 sentences):** How did changing the null value change the conclusion? What does this tell you about the relationship between the null hypothesis and the observed estimate?

> *Changing my null value from 12 to 10 did change the conclusion. When the null value was 12, it was inside the 95% CI and produced a large p-value (0.169), so it failed to reject $H_0$. When we changed the null value to 10, the z_stat_alt was large and the p-value became very small (0.0005), so here we can reject $H_0$. This shows the null value is far from the observed estimate and the data looks surprising. *

---

# Part II — Comparing group means with t-tests

In Part I, you tested a claim about **one population mean**.

Now you will extend the same logic to compare **two group means**.

The key idea is:

> A two-sample t-test asks whether the difference between two group means is large relative to the uncertainty in that difference.

You will first work with a built-in grouping variable (**day vs night**), then choose a second grouping variable of your own.

---

## 11) Create a day/night grouping variable

If you already created `day_night` earlier, you can reuse it. Otherwise, create it here.

```{r make_daynight, echo=TRUE, eval=TRUE}
# GOAL: create a day/night grouping variable from `daytime`.

df <- df %>%
  mutate(
    day_night = if_else(daytime == 1, "Day", "Night")
  ) %>% 
  filter(!is.na(daytime))

df %>% count(day_night)
```

---

## 12) Choose a response variable for the t-test

Pick one numeric response variable to compare between groups.

Possible examples:
- `ta`
- `swc`
- `fc`
- `gpp`
- `le`

```{r choose_response_ttest, echo=TRUE, eval=TRUE}
response_var <- "swc"

# CHECK: does this variable exist?
response_var %in% names(df)
```

---

## 13) Visualize the response by day/night

Create a plot that shows how your response variable differs between Day and Night.

You may use:
- `geom_boxplot()`
- `geom_violin()`
- `geom_jitter()`
- or a combination

```{r plot_daynight, echo=TRUE, eval=TRUE}
# GOAL: visualize the response by day/night group.

df %>%
  filter(!is.na(.data[[response_var]])) %>%
  ggplot(aes(x = day_night, y = .data[[response_var]], fill = day_night)) +
  geom_boxplot(alpha = 0.8) +
  theme_classic(base_size = 18) +
  labs(
    x = NULL,
    y = response_var,
    title = "Response by day vs night"
  )
```

**Prompt (2–3 sentences):** Before running the test, what do you predict?  
Do the groups look different in center? In spread?

> *Before running the test I would predict that soil water content will not differ that much. The box plot shows that the groups have similar centers and also the spread looks similar between the groups.*

---

## 14) Summarize the two groups

Compute the group means, SDs, and sample sizes.

```{r summarize_daynight, echo=TRUE, eval=TRUE}
# GOAL: summarize the response variable by day/night group.

group_summary <- df %>%
  filter(!is.na(.data[[response_var]])) %>%
  group_by(day_night) %>%
  summarise(
    mean_response = mean(.data[[response_var]], na.rm = TRUE),
    sd_response = sd(.data[[response_var]], na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

group_summary
```

---

## 15) State the hypotheses for the day/night comparison

Write hypotheses in words and symbols.

Use this structure:

- Null: the two group means are equal
- Alternative: the two group means are different

Fill in:

> My response variable is `soil water content (swc)`

> $H_0$: `The mean soil water temperature is the same between Day and Night.`

> $H_A$: `The mean soil water temperature isn't the same between Day and Night. `

---

## 16) Run a two-sample t-test

Use `t.test()` to compare the means of your response variable between Day and Night.

```{r ttest_daynight, echo=TRUE, eval=TRUE}
# GOAL: run a two-sample t-test by formula.

t.test(
  swc ~ day_night,
  data = df
)
```

**Prompt (4–6 sentences):** Report the p-value and confidence interval.  
Would you reject the null hypothesis at $\alpha = 0.05$?  
What does the result imply about day vs night differences in your chosen variable?

> *The p-value is < 2.2e-16 which is small. The 95% CI interval is -1.399 to -0.90, which doesn't include 0. Because the p-value is extremely small and the confidence interval excludes zero, we reject the null hypothesis that the two group means are equal. This result implies that soil water content differs significantly between day and night. Also the mean SWC is lower during the day (35.38) than at night (36.52).*

---

## 17) Connect the t-test to a confidence interval

Extract the confidence interval and interpret it as a difference in means.

```{r ttest_store, echo=TRUE, eval=TRUE}
# GOAL: store the t-test result in an object.

tt_daynight <- t.test(
  swc ~ day_night,
  data = df
)

tt_daynight$conf.int
```

**Prompt (2–3 sentences):** Does the confidence interval include 0?  
How does that connect to the hypothesis test decision?

> *No, the confidence interval doesn't include 0. Because the interval excludes 0, this supports the decision to reject the null hypothesis.*

---

## 18) Bootstrap view of the same comparison (conceptual extension)

Create bootstrap distributions of the **group means** for Day and Night separately.

```{r bootstrap_group_means, echo=TRUE, eval=TRUE}
set.seed(100)

B <- 2000   # e.g., 2000

x_day <- df %>%
  filter(day_night == "Day") %>%
  pull(.data[[response_var]]) %>%
  na.omit()

x_night <- df %>%
  filter(day_night == "Night") %>%
  pull(.data[[response_var]]) %>%
  na.omit()

boot_day <- replicate(
  B,
  mean(sample(x_day, size = length(x_day), replace = TRUE))
)

boot_night <- replicate(
  B,
  mean(sample(x_night, size = length(x_night), replace = TRUE))
)

boot_groups <- bind_rows(
  tibble(group = "Day", mean_est = boot_day),
  tibble(group = "Night", mean_est = boot_night)
)
```

```{r plot_boot_groups, echo=TRUE, eval=TRUE}
# GOAL: visualize the bootstrap distributions by group.

ggplot(boot_groups, aes(x = mean_est, fill = group)) +
  geom_histogram(alpha = 0.45, bins = 50, position = "identity") +
  theme_classic(base_size = 18) +
  labs(
    x = paste0("Bootstrap mean of ", response_var),
    y = "Count",
    title = "Bootstrap distributions of group means"
  )
```

**Prompt (3–4 sentences):** How does this bootstrap visualization help you understand the t-test result?  
Do the plausible group means strongly overlap, or are they clearly separated?

> *The bootstrap visualization shows the two distributions are clearly separated with little to no overlap. This shows that the values for the group means are different. The visual itself explains why the t-test produced a small p-value and rejected the null hypothesis. Because the plausible means for Day and Night do not overlap much, it provides strong evidence that soil water content differs between the two groups.*

---

## 19) Choose a second grouping variable of your own

Now define **one additional grouping variable** and repeat the t-test workflow.

Possible ideas:
- month grouped into two seasons (e.g., winter vs summer)
- a high/low split of another continuous variable
- a wet/dry split of SWC
- a daytime subset grouped by month category

You must:
- define the grouping variable clearly
- explain why the grouping is scientifically meaningful

```{r choose_second_group, echo=TRUE, eval=TRUE}
# GOAL: create your own two-level grouping variable.

df <- df %>%
  mutate(
    my_group = if_else(month %in% c(6,7,8), "Summer", "Other")
  ) %>% 
  filter(!is.na(my_group))

df %>% count(my_group)

```

**Prompt (2–3 sentences):** Why did you choose this grouping variable? What scientific contrast are you testing?

> *I chose this grouping variable into summer vs non-summer months because seasonal changes can strongly influence ecosystem processes for example such as ta, swc, and etc. This grouping allows us to test whether the response variable differs between warmer summer months and the rest of the year. *

---

## 20) Visualize and test your second grouping variable

Repeat the workflow:
1. visualize
2. summarize
3. state hypotheses
4. run `t.test()`
5. interpret

```{r visualize_second_group, echo=TRUE, eval=TRUE}
# GOAL: visualize your response by your chosen grouping variable.
df %>%
  ggplot(aes(x = my_group, y = ta, fill = my_group)) +
  geom_boxplot(alpha = 0.8) +
  theme_classic(base_size = 18) +
  labs(
    x = NULL,
    y = "le",
    title = "Not Summer"
  )
```

```{r summarize_second_group, echo=TRUE, eval=TRUE}
# GOAL: summarize your response variable by your chosen grouping variable.

group_variable_summary <- df %>%
  filter(!is.na(ta)) %>%
  group_by(my_group) %>%
  summarise(
    mean_response = mean(ta, na.rm = TRUE),
    sd_response = sd(ta, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

group_variable_summary
```

```{r ttest_second_group, echo=TRUE, eval=TRUE}
# GOAL: run a two-sample t-test for your chosen grouping variable.
tt_notsummer <- t.test(
  ta ~ day_night,
  data = df
)
```

**Prompt (5–6 sentences):** Compare the day/night t-test to your second grouping-variable test.  
Which contrast was stronger? Which was more scientifically meaningful?  
Did the two tests produce similar or different conclusions?

> *The day/night t-test showed a very strong difference, with a very small p-value and a large difference in the group means. The second grouping-variable test was even stronger, with a very small p-value (8.94 e-299) and a CI between 5.57-6.18, indicating a large separation between the groups. Scientifically, the day/night difference is meaningful because many environmental variables change over the daily cycle due to sunlight and energy exchange. Both tests led to the same conclusion and thus we reject the null hypothesis, meaning the group means are significantly different in each case.*

---

## 21) Final reflection

Answer in **4–6 sentences**:

- How is a two-sample t-test similar to the one-mean hypothesis test from Part I?
- What changes when we move from “testing one mean” to “testing a difference in means”?
- How did the bootstrap group-mean plot help your understanding?

> *A two-sample t-test is similar to a one-mean hypothesis test because both compare an observed estimate to what we would expect from a null hypothesis. We then use a test statistic and p-value to evaluate the data set. In the one-mean test, we compare a single sample mean to a hypothesized value, while in the two-sample test we compare the means of two groups to see if they're different. The difference is that the uncertainty now comes from both groups, so we can estimate the standard error of the difference between the two means rather than just one mean. The bootstrap group-mean plot helped visualize the sampling distributions of the group means and showed how much they overlap and seeing the distributions made it easier to understand why the t-test concluded that the group means were different.*

---

# Submission

- Knit your `.Rmd` to HTML.
- Commit and push both the `.Rmd` and knitted `.html` to your GitHub repo.
- Submit the GitHub repository link on Canvas.
